Lambda Calculus through Tropical Mathematics

### 2023-01-12 section3B.tex
Ho aggiustato delle cose in Sezione III
In particolare, ho introdotto lì il famoso esempio

\begin{remark}[Tropical Laurent series]
 As usual, a matrix $t\in\HOM{\LREL_!}{X}{Y}$ operates as a linear map $\widehat t:\Lawv^{!X}\to\Lawv^Y$.
 But we can also ``express it in the base $X$'', i.e.\ see it as a map $t^!:\Lawv^X\to\Lawv^Y$, by setting $t^!(x):=t\circ_! x$ (we are identifying $\Lawv^X$ with the set $\HOM{\LREL_!}{\emptyset}{X}$ of the \emph{points} of $X$).
 This is the notion of \emph{non-linear} map generated by the CCC-structure of $\LREL_!$.
 Concretely, we have:
 $t^!(x)_b=\inf\limits_{\mu\in !X} \set{\mu x+ t_{\mu,b}}$, where $\mu x:=\sum\limits_{a\in X} \mu(a)x_a$.
 Since in the general case of $\QREL$, $t^!$ is a Laurent series with operations in $Q$, let us call \emph{tropical Laurent series} the functions of shape $t^!$ for some $t\in\HOM{\LREL_!}{X}{Y}$.
\end{remark}

>>>>

\begin{remark}[Tropical Laurent series]
 As usual, a matrix $t\in\HOM{\LREL_!}{X}{Y}$ operates as a linear map $\widehat t:\Lawv^{!X}\to\Lawv^Y$.
 But we can also ``express it in the base $X$'', i.e.\ see it as a map $t^!:\Lawv^X\to\Lawv^Y$, by setting $t^!(x):=t\circ_! x$ (we are identifying $\Lawv^X$ with the set $\HOM{\LREL_!}{\emptyset}{X}$ of the \emph{points} of $X$).
 This is the notion of \emph{non-linear} map generated by the CCC-structure of $\LREL_!$.
 Concretely, we have:
 \[t^!(x)_b=\inf\limits_{\mu\in !X} \set{\mu x+ t_{\mu,b}}\] where $\mu x:=\sum\limits_{a\in X} \mu(a)x_a$.
 Since in the general case of $\QREL$, $t^!$ is a Laurent series with operations in $Q$, let us call \emph{tropical Laurent series} the functions of shape $t^!$ for some $t\in\HOM{\LREL_!}{X}{Y}$.
\end{remark}

\begin{remark}
 Indentifying $!\set{*}\simeq \N$ and $\Lawv^{\set{*}}\simeq\Lawv$, the tropical Laurent series generated by the morphisms in $\HOM{\LREL_!}{\set{*}}{\set{*}}$ are exactly the functions $f:\Lawv\to\Lawv$ of shape $f(x)=\inf\limits_{n\in\N}\set{nx+\widehat f(n)}$, for some $\widehat f:\N\to\Lawv$.
Remark that we find usual \emph{tropical polynomials} of tropical geometry as a particular case: they corrispond to the ones for which the support $\set{n\in\N\mid\widehat f(n)\neq\infty}$ of $\widehat f$ is \emph{finite}.
\end{remark}

\begin{proposition}\label{prop:descrete}
 The interpretation $\model M$ of a $\lam$-term $M$ in $\LREL$ is a \emph{descrete} matrix, i.e.\ its coefficients are either $0$ or $\infty$.
\end{proposition}
\begin{proof}
 First easily prove that composition preserves descreteness.
 Then go by straightforward induction on $M$, using that the projections and evaluation of  $\LREL$ are descrete.
\end{proof}

\begin{example}
 The function $f:\Lawv\to\Lawv$, $f(x):=\inf\limits_{n\in\N}\set{nx+\frac{1}{2^n}}$ is a tropical Laurent series: it is of shape $f=t^!$, for $t\in\Lawv^{!\set{*}\times\set{*}}$, $t_{\mu,*}:=2^{-\# \mu}$.
By Proposition~\ref{prop:descrete}, $f$ is not the interpretation of a $\lam$-term, because its matrix is not descrete. Therefore $\LREL_!$ is not a full-complete model of $\STLC$.
\end{example}

### 2023-01-12 section3C.tex
Ho aggiustato delle cose in Sezione III
In particolare, ho introdotto lì il famoso esempio

Let us show the differential structure of $\LREL_!$ (remember that the Cartesian product of $\LREL_!$ is the disjoint union $+$): 
\begin{definition}
 The \emph{tropical differential operator} is the map $D:\HOM{\LREL}{!X}{Y}\to \HOM{\LREL}{!(X+X)}{Y}$,
 $(Dt)_{\mu_1\oplus\mu_2,b}:=t_{\mu_2*\mu_1,b}$ if $\#\mu_1=1$ and $(Dt)_{\mu_1\oplus\mu_2,b}:=\infty$ otherwise.
\end{definition}

\begin{example}
 \`E BENE METTERE UN ESEMPIO DI COME OPERA QUESTO DIFFERENZIALE (senn\`o non si capisce).
 Secondo me conviene con un $D^2$ di qualcosa.
\end{example}

\begin{theorem}\label{thm:LREL!CCDC}
 $\LREL_!$ equipped with $D$, is a CC$\partial$C.
\end{theorem}
\begin{proof}
 The properties 1) and 2) are tedious technical checks.
 If the first are quite straightforward, while the second deserve more attention.
 Let us give the idea for ?
\end{proof}

@@

\begin{theorem}\label{thm:modelsTaylor}
 In $\LREL_!$ equipped with $D$, all morphisms can be Taylor expanded, i.e.\ in $\LREL_!$ we have the following equality:
 \[
  \RM{eval}^{X,Y}\circ_!\langle t,s\rangle =
  \inf\limits_{n\in\N}
  \set{((\dots((\Lambda^- t)\star s)\star \dots)\star s)\circ \langle \RM{id},\infty \rangle}.
 \]
 for all $t\in\HOM{\LREL_!}{Z}{X\multimap Y}$, $s\in\HOM{\LREL_!}{Z}{X}$.
\end{theorem}
Here, $\Lambda^-:\HOM{\LREL}{!Z}{X\multimap Y}\to \HOM{\LREL}{!(Z+X)}{Y}$ is the uncarry operator and $\star:\HOM{\LREL}{!(Z+X)}{Y}\times\HOM{\LREL}{!Z}{X}\to \HOM{\LREL}{!(Z+X)}{Y}$ is an operator corresponding to the syntactical \emph{differential substitution}.
\begin{proof}
 Che ci scriviamo ??
\end{proof}

@@

- Discuss the tropical Taylor expansion $\trop\C T(M)$ ??

>>>>

Let us show the differential structure of $\LREL_!$ (remember that the Cartesian product of $\LREL_!$ is the disjoint union $+$):

\begin{definition}
 Let $\widetilde{(\cdot)}:\HOM{\LREL}{!X}{Y}\to \HOM{\LREL}{!(X+X)}{Y}$, $\widetilde t_{\mu\oplus\rho,b}:= t_{\mu*\rho,b}$, and $d:\HOM{\LREL}{!(X+X)}{Y}\to \HOM{\LREL}{!(X+X)}{Y}$, $(dt)_{\mu\oplus\rho,b}:=t_{\rho\oplus\mu,b}$ if $\#\mu_1=1$ and $(dt)_{\mu\oplus\rho,b}:=\infty$ otherwise.
 The \emph{tropical differential operator} is the map $D:=d\circ\widetilde{(\cdot)}:\HOM{\LREL}{!X}{Y}\to \HOM{\LREL}{!(X+X)}{Y}$.
\end{definition}

\begin{example}
For $t\in\HOM{\LREL}{!X}{Y}$, we have:
 $(Dt)_{\mu\oplus\rho,b}=t_{\rho*\mu,b}$ if $\#\mu_1=1$ and it is $\infty$ otherwise.
 Analogously, we have:
 $D^2 t\in\HOM{\LREL_!}{(X+X)+(X+X)}{Y}$, where $(D^2 t)_{(\rho\oplus\rho')\oplus(\nu\oplus\nu'),b}$ equals $t_{\nu*\nu'*\rho',b}$ if $\rho=\emptyset$ and $\#\rho'=1=\#\nu$; it equals $t_{\rho*\nu',b}$ if $\rho'=\emptyset=\nu$ and $\#\rho=1$; it equals $\infty$ otherwise.
\end{example}

\begin{theorem}\label{thm:LREL!CCDC}
 $\LREL_!$ equipped with $D$, is a CC$\partial$C.
\end{theorem}
\begin{proof}
 The properties 1) are quite straightforward.
 Properties 2) are tedious technical checks.
\end{proof}

@@

\begin{theorem}\label{thm:modelsTaylor}
 In $\LREL_!$ equipped with $D$, all morphisms can be Taylor expanded, i.e.\ in $\LREL_!$ we have the following equality:
 \[
  \RM{eval}^{X,Y}\circ_!\langle t,s\rangle =
  \inf\limits_{n\in\N}
  \set{((\dots((\Lambda^- t)\star s)\star \dots)\star s)\circ_! \langle \RM{id},\infty \rangle}.
 \]
 for all $t\in\HOM{\LREL_!}{Z}{X\multimap Y}$, $s\in\HOM{\LREL_!}{Z}{X}$.
\end{theorem}
Here, $\Lambda^-:\HOM{\LREL}{!Z}{X\multimap Y}\to \HOM{\LREL}{!(Z+X)}{Y}$ is the uncarry operator and $\star:\HOM{\LREL}{!(Z+X)}{Y}\times\HOM{\LREL}{!Z}{X}\to \HOM{\LREL}{!(Z+X)}{Y}$ is an operator corresponding to the syntactical \emph{differential substitution}.
\begin{proof}
 Che ci scriviamo ??
\end{proof}

@@
